### 2018-03-15 submission/manuscript.Rmd
fix capitalization error
complete adding references

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases [ref]. For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [refs]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus Influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

>>>>

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can increase the difference between these types of 16S rRNA genes and over-emphasizing the actual difference [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

### 2018-03-15 submission/manuscript.tex
fix capitalization error
complete adding references

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community {[}ref{]}. Similarily, this overgrowth could still
occur if the preservation media does not adequately inhibit growth
{[}ref{]}. Reports have also shown that changes in specific community
members might occur due to differing susceptibility to freeze thaw
cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
has been shown to add community members and the contribution of these
contaminant members grows larger with lower biomass samples (Salter et
al., 2014). Recent studies have shown that the majority of these biases
due to either preservation or extraction tend to be smaller than the
overall biological signal being measured {[}refs{]}. However, the
contribution of PCR bias to this overall workflow is not well
characterized since these studies use the same PCR approach while
varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results
{[}refs{]}. Many of these sources of biases are made worse as cycle
number increases {[}ref{]}. For example, the selective amplification of
the dominant sequence can increase the difference between the most and
next abundant sequence, over-emphasizing the actual difference
{[}ref{]}. Both amplification error and non-specific amplification
(e.g.~incorrect amplicon size products) can also increase as cycle
number increases which can drastically change commonly used diversity
measures {[}refs{]}. Additionally, chimeras can form from an aborted
extension followed by subsequent priming error and secondary extension
and even if it has not been explicitly shown to increase with cycle
number it is easy to postulate that this does happen and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus Influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) {[}refs{]}. Additionally, there are
multiple families of DNA polymerases that have their own error rate and
proof reading capcity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
results have not been well studied like some of the other previously
mentioned sources of PCR-based bias.

@@

\hypertarget{refs}{}
\hypertarget{ref-benjamini_controlling_1995}{}
Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. \emph{Journal of
the Royal Statistical Society. Series B (Methodological)} 57:289--300.

@@

@@

@@

@@

@@

@@

@@

>>>>

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community (Amir et al., 2017). Similarily, this overgrowth
could still occur if the preservation media does not adequately inhibit
growth (Song et al., 2016; Luo et al., 2016). Reports have also shown
that changes in specific community members might occur due to differing
susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
2015). Additionally, reagent contamination has been shown to add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Recent
studies have shown that the majority of these biases due to either
preservation or extraction tend to be smaller than the overall
biological signal being measured (Song et al., 2016; Bassis et al.,
2017). However, the contribution of PCR bias to this overall workflow is
not well characterized since these studies use the same PCR approach
while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can increase the
difference between these types of 16S rRNA genes and over-emphasizing
the actual difference (Polz \& Cavanaugh, 1998). Both amplification
error and non-specific amplification (e.g.~incorrect amplicon size
products) can also increase as cycle number increases which can
drastically change commonly used diversity measures (Acinas et al.,
2005; Santos et al., 2016). Additionally, chimeras can form from an
aborted extension followed by subsequent priming error and secondary
extension and even if it has not been explicitly shown to increase with
cycle number it is easy to postulate that this does happen and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capcity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed results have not been well studied
like some of the other previously mentioned sources of PCR-based bias.

@@

\hypertarget{refs}{}
\hypertarget{ref-Acinas2005}{}
Acinas SG., Sarma-Rupavtarm R., Klepac-Ceraj V., Polz MF. 2005.
PCR-induced sequence artifacts and bias: Insights from comparison of two
16S rRNA clone libraries constructed from the same sample. \emph{Applied
and Environmental Microbiology} 71:8966--8969. DOI:
\href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.

\hypertarget{ref-Amir2017}{}
Amir A., McDonald D., Navas-Molina JA., Debelius J., Morton JT., Hyde
E., Robbins-Pianka A., Knight R. 2017. Correcting for microbial blooms
in fecal samples during room-temperature shipping. \emph{mSystems}
2:e00199--16. DOI:
\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.

\hypertarget{ref-storage_Bassis_2017}{}
Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
samples and storage conditions on bacterial community profiles.
\emph{BMC Microbiology} 17. DOI:
\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.

\hypertarget{ref-benjamini_controlling_1995}{}
Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. \emph{Journal of
the Royal Statistical Society. Series B (Methodological)} 57:289--300.

\hypertarget{ref-Burkardt2000}{}
Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.

@@

\hypertarget{ref-Eckert1991}{}
Eckert KA., Kunkel TA. 1991. DNA polymerase fidelity and the polymerase
chain reaction. \emph{Genome Research} 1:17--24. DOI:
\href{https://doi.org/10.1101/gr.1.1.17}{10.1101/gr.1.1.17}.

@@

\hypertarget{ref-Gorzelak2015}{}
Gorzelak MA., Gill SK., Tasnim N., Ahmadi-Vand Z., Jay M., Gibson DL.
2015. Methods for improving human gut microbiome data by reducing
variability through sample processing and storage of stool. \emph{PLOS
ONE} 10:e0134802. DOI:
\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.

@@

\hypertarget{ref-Kebschull2015}{}
Kebschull JM., Zador AM. 2015. Sources of PCR-induced distortions in
high-throughput sequencing data sets. \emph{Nucleic Acids
Research}:gkv717. DOI:
\href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.

@@

\hypertarget{ref-preservation_Luo_2016}{}
Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
2016. Effects of specimen collection methodologies and storage
conditions on the short-term stability of oral microbiome taxonomy.
\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.

\hypertarget{ref-Meisel2016}{}
Meisel JS., Hannigan GD., Tyldsley AS., SanMiguel AJ., Hodkinson BP.,
Zheng Q., Grice EA. 2016. Skin microbiome surveys are strongly
influenced by experimental design. \emph{Journal of Investigative
Dermatology} 136:947--956. DOI:
\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.

@@

\hypertarget{ref-polz_bias_1998}{}
Polz MF., Cavanaugh CM. 1998. Bias in template-to-product ratios in
multitemplate PCR. \emph{Applied and Environmental Microbiology}
64:3724--3730.

@@

\hypertarget{ref-BautistadelosSantos2016}{}
Santos QMB-d los., Schroeder JL., Blakemore O., Moses J., Haffey M.,
Sloan W., Pinto AJ. 2016. The impact of sampling, PCR, and sequencing
replication on discerning changes in drinking water bacterial community
over diurnal time-scales. \emph{Water Research} 90:216--224. DOI:
\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.

@@

\hypertarget{ref-preservation_Song_2016}{}
Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
R. 2016. Preservation methods differ in fecal microbiome stability,
affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
DOI:
\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.

\hypertarget{ref-Sze2015}{}
Sze MA., Dimitriu PA., Suzuki M., McDonough JE., Campbell JD., Brothers
JF., Erb-Downward JR., Huffnagle GB., Hayashi S., Elliott WM., Cooper
J., Sin DD., Lenburg ME., Spira A., Mohn WW., Hogg JC. 2015. Host
response to the lung microbiome in chronic obstructive pulmonary
disease. \emph{American Journal of Respiratory and Critical Care
Medicine} 192:438--445. DOI:
\href{https://doi.org/10.1164/rccm.201502-0223oc}{10.1164/rccm.201502-0223oc}.

\hypertarget{ref-Wang1996}{}
Wang GCY., Wang Y. 1996. The frequency of chimeric molecules as a
consequence of PCR co-amplification of 16S rRNA genes from different
bacterial species. \emph{Microbiology} 142:1107--1114. DOI:
\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
